---
title: 'Final Project 2: Reproducible Report on COVID19 Data'
author: "J. Suarez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
```

### From Class:
```{r insertData}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_confirmed_global.csv")

urls <- str_c(url_in, file_names)
urls
```
```{r import_data, message = FALSE}
global_cases <- read_csv(urls[4])
global_deaths <- read_csv(urls[3])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[2])
```

```{r tidy_global_data}
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat, Long))
```
```{r tidy_global_deaths}
global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat, Long))

global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>% mutate(date = mdy(date))

global <- global %>% filter(cases>0)
```
```{r tidy_us_cases}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
```

```{r tidy_us_deaths}
US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
```

```{r combine_US}
US <- US_cases %>% full_join(US_deaths)
```

```{r add_pop_to_global}
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url)

global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key.x)
```
```{r organize_US_by_state}
US_by_state <- US %>% 
  group_by(Province_State, Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = (sum(Population))) %>% 
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, 
         cases, deaths, deaths_per_mill, Population) %>% 
  ungroup()
```

```{r organize_US_totals}
US_totals <- US_by_state %>% 
  group_by(Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = (sum(Population))) %>% 
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, 
         cases, deaths, deaths_per_mill, Population) %>% 
  ungroup()
```

### Individual Visualizations, Model, and Analysis

```{r organize_mexico_data}
  mexico_data <- global %>% filter(Country_Region == "Mexico") %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% mutate(deaths_per_thou = deaths * 1000 / Population) %>% mutate(cases_per_mill = cases * 1000000 / Population) %>% mutate(cases_per_thou = cases * 1000 / Population)

  mexico_data
```

```{r organize_canada_data}


canada_data <- global %>% filter(Country_Region == "Canada") 

summary(canada_data)

canada_data <- canada_data %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases, na.rm = TRUE), deaths = sum(deaths, na.rm = TRUE), 
            Population = (sum(Population, na.rm = TRUE)))


canada_data <- canada_data %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% mutate(deaths_per_thou = deaths * 1000 / Population) %>% mutate(cases_per_mill = cases * 1000000 / Population) %>% mutate(cases_per_thou = cases * 1000 / Population)
  
canada_data

 
```

```{r add_to_US_state_data}
US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths)) %>% mutate(deaths_per_thou = deaths * 1000 / Population) %>% mutate(cases_per_mill = cases * 1000000 / Population) %>% mutate(cases_per_thou = cases * 1000 / Population)
  
US_totals

```

```{r plot_US_data, message = FALSE}

#total cases and deaths
US_totals %>% filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line() +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y = "NULL")

#new cases and deaths
US_totals %>%  ggplot(aes(x = date, y = new_cases)) +
  geom_line() +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New Incidents of COVID19 in US", y = "NULL")

```

```{r plot_mexico_data, message = FALSE}
#total cases and deaths
mexico_data %>% filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line() +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in Mexico", y = "NULL")

#new cases and deaths
mexico_data %>%  ggplot(aes(x = date, y = new_cases)) +
  geom_line() +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New Incidents of COVID19 in Mexico", y = "NULL")

```

```{r plot_canada_data, message = FALSE}
#total cases and deaths
canada_data %>% filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line() +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in Canada", y = "NULL")

#new cases and deaths
canada_data %>%  ggplot(aes(x = date, y = new_cases)) +
  geom_line() +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New Incidents of COVID19 in Canada", y = "NULL")

```

```{r model_US_data, message = FALSE}
US_mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_totals)
summary(US_mod)

US_totals %>% mutate(pred = predict(US_mod))

US_tot_w_pred <- US_totals %>% mutate(pred = predict(US_mod))
US_tot_w_pred
US_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") + labs(title = str_c("Linear Regression Model of US COVID Incidents"))
```


```{r model_mexico_data}
mex_mod <- lm(deaths_per_thou ~ cases_per_thou, data = mexico_data)
summary(mex_mod)

mexico_data %>% mutate(pred = predict(mex_mod))
mexico_w_pred <- mexico_data %>% mutate(pred = predict(mex_mod))
mexico_w_pred
mexico_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") + labs(title = str_c("Linear Regression Model of Mexico COVID Incidents"))
```

```{r model_canada_data}
can_mod <- lm(deaths_per_thou ~ cases_per_thou, data = canada_data)
summary(can_mod)

canada_data %>% mutate(pred = predict(can_mod))
canada_w_pred <- canada_data %>% mutate(pred = predict(can_mod))
canada_w_pred
canada_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") + labs(title = str_c("Linear Regression Model of Canada COVID Incidents"))
```

#### Analysis of Data and Conclusions

After looking at all of the data, it appears that Canada was the country to recover the least quickly despite having the smallest population of all the countries in North America. I base this assumption off the model visualization where Canada was the only country whose relationship between deaths per thousand and cases per thousand kept increasing even beyond the predicted linear model. Looking at total cases and deaths in all three countries, it comes to no surprise the United States had the most in both categories due to their large population when compared to Mexico and Canada. The models for each country seem to show that cases per thousand are a good indicator of deaths per thousand for the three countries. There are some outliers which may be caused by additional variables. 

Questions arise on what particularly about these three countries, while being neighbors, have largely different experiences with COVID. While most is potentially due to population and population density (of which more data would be useful for further analysis), there may be other factors such as the level of government intervention, availability of resources, and human behavior may have had an additional impact. 

There may be some bias present due to the way the data was collected. Each of the three countries have separate laws which dictate how to record and report data. There may be cases or deaths in any of the data sets that went unreported. There may be even cases or deaths reported that should not be attributed to COVID due to false negatives or other types of errors.  

